Local Type Inference for Polarised System F with Existentials

### 2023-10-31 oopsla24/tex/03-declarative-system.tex
WIP: Algorithm
Finish the Algorithmic section
Corrections

\begin{figure}[h]
  \begin{multicols}{2}
    \ottdefnDOneNsub{}

    \ottdefnDOnePsup{}
  \end{multicols}
  \hfill\\
  \begin{multicols}{2}
    \ottdefnDOneNeq{}

    \ottdefnDOnePeq{}
  \end{multicols}
  \caption{Declarative Subtyping}
  \label{fig:declarative-subtyping}
\end{figure}

@@

\subsubsection{Properties of the Declarative Subtyping}
A property that we extensively use is that the subtyping is reflexive and transitive,
and agrees with substitution.

@@

\begin{figure}[h]
  \label{alg:type-nf}
  \hfill

  \begin{multicols}{2}
  \ottdefnNrmNNorm{}
  \\
  $[[ord varset in iN]]$ returns a list of variables
  $[[varset ∩ fv(iN)]]$ in the order of their first occurrence in $[[iN]]$
  \columnbreak\\
  \ottdefnNrmPNorm{}
  \\
  $[[ord varset in iP]]$ returns a list of variables
  $[[varset ∩ fv(iP)]]$ in the order of their first occurrence in $[[iP]]$
  \end{multicols}

@@

  \begin{multicols}{2}
  \ottdefnDTNInf{}
  \columnbreak\\
  \ottdefnDTPInf{}
  \ottdefnDTSpinInf{}
  \end{multicols}
  \hfill

  \caption{Declarative Subtyping}
  \label{fig:declarative-inference}
\end{figure}

>>>>

\begin{figure}[h]
  \begin{multicols}{2}
    \ottdefnDOneNsubLabeled{}

    \ottdefnDOnePsupLabeled{}
  \end{multicols}
  \hfill\\
  \begin{multicols}{2}
    \ottdefnDOneNeqLabeled{}

    \ottdefnDOnePeqLabeled{}
  \end{multicols}
  \caption{Declarative Subtyping}
  \label{fig:declarative-subtyping}
\end{figure}

@@

\subsubsection{Properties of the Declarative Subtyping}
\label{sec:decl-subtyping-properties}

A property that is important for the subtyping algorithm, 
in particular for the type \emph{upgrade} procedure (\cref{todo}),
is the preservation of free variables by subtyping.
Informally, it says that the free variables
of a positive type cannot disappear in its subtypes,
and the free variables of a negative type
cannot disappear in its supertypes.

\begin{property}[Subtyping Preserves Free Variables]
  \label{prop:subtyping-preserves-fv}
  Let us assume that all the mentioned types are well-formed in $[[Γ]]$. Then
  $[[Γ ⊢ iN1 ≤ iN2]]$ implies $[[fv(iN1) ⊆ fv(iN2)]]$,
  and $[[Γ ⊢ iP1 ≥ iP2]]$ implies $[[fv(iP1) ⊆ fv(iP2)]]$.
\end{property}

Another property that we extensively use is that the subtyping is reflexive and transitive,
and agrees with substitution.

@@

  \begin{multicols}{2}
  \ottdefnNrmNNormLabeled{}
  \\
  $[[ord varset in iN]]$ returns a list of variables
  $[[varset ∩ fv(iN)]]$ in the order of their first occurrence in $[[iN]]$
  \columnbreak\\
  \ottdefnNrmPNormLabeled{}
  \\
  $[[ord varset in iP]]$ returns a list of variables
  $[[varset ∩ fv(iP)]]$ in the order of their first occurrence in $[[iP]]$
  \end{multicols}

@@

\begin{figure}[h]
  \ottdefnsDTLabeled
  \caption{Declarative Inference}
  \label{fig:declarative-inference}
\end{figure}

### 2023-10-31 oopsla24/tex/04-algorithm.tex
Corrections

        % - Justify Adding α̂± 
        % - algorithmic type contexts Ξ and constraint context Θ
        % - Describe how wf, nf,... are changed
        % - Constraints -- important part of the algorithm

        % - Dependency Graph

In this section, we present the algorithmization of the system described before
Shadowing the declarative system, the algorithm has two main parts: 
the subtyping and the type inference, which we discuss in this section one after another.

@@

\begin{description}
        \item[Positive Algorithmic Variables] $[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, \dots 
        \item[Negative Algorithmic Variables] $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, \dots
        \item[Positive Algorithmic Types] $[[uP]] = \dots \mid [[α̂⁺]]$  
        \item[Negative Algorithmic Types] $[[uN]] = \dots \mid [[α̂⁻]]$
        \item[Algorithmic Type Context] $[[Ξ]] = \{[[α1̂±]], \dots, [[αn̂±]]\}$
        where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct 
\end{description}

\paragraph{Algorithmic Variables}
Both subtyping and the inference algorithms are represented by sets of inference rules, 
to simplify the soundness and completeness proofs \wrt the declarative specification. 
The terms these rules manipulate we call \emph{algorithmic}. They extend the 
previously defined declarative terms and types by adding \emph{algorithmic type variables}
(\aka unification variables). The algorithmic variables represent unknown types, 
which cannot be inferred immediately but are promised to be instantiated
as the algorithm proceeds.

We denote algorithmic variables as $[[α̂⁺]]$, $[[β̂⁻]]$, \dots to distinguish them from
normal variables $[[α]]$, $[[β]]$, \dots and also to smooth out the transition
from the declarative to the algorithmic system, when we replace the quantified variables $[[pas]]$ 
with their algorithmic counterpart $[[puas]]$.
The procedure of replacing declarative variables with algorithmic ones we call 
\emph{algorithmization} and denote as $[[ nuas/nas ]]$ and $[[ puas/pas ]]$.

\paragraph{Algorithmic Types}
The syntax of  algorithmic types is the syntax of declarative types extended
with algorithmic type variables: we add positive algorithmic variables $[[α̂⁺]]$ 
to the positive types, and negative algorithmic variables $[[α̂⁻]]$ to the negative types.
Notice that these variables cannot be abstracted by the quantifiers $[[∀]]$ and $[[∃]]$.
We denote the algorithmic types

\paragraph{Algorithmic Contexts and Well-formedness}
To specify when algorithmic types are well-formed, 
we define algorithmic contexts $[[Ξ]]$ as sets of algorithmic variables.
Then $[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness
judgment of algorithmic terms defined as expected: in addition to the declarative definition
\cref{todo}, we also check that each algorithmic variable is in the context $[[Ξ]]$.
\begin{figure}[h]
\begin{multicols}{2}
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleWFATPUVar{}}
  \columnbreak\\
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleWFATNUVar{}}
\end{multicols}
\label{fig:algo-wf}
\caption{Well-formedness of Algorithmic Types}
\end{figure}

\paragraph{Algorithmic Normalization}
Similarly to well-formedness, the normalization of algorithmic types is defined
by extending the declarative definition trivially with the algorithmic variables.

@@

\subsection{Type Constraints}
As the algorithm proceeds, it accumulates the information 
about the algorithmic type variables in the form of \emph{constraints}.
In our system, the constraints can be of two kinds: 
\emph{subtyping constraints} and \emph{unification constraints}.
The subtyping constraint can only have a positive shape $[[α̂⁺ :≥ iP]]$, \ie it 
restricts a positive algorithmic variable to be a 
supertype of a certain declarative type---this is one of the invariants that we 
preserve in the algorithm.
The unification constraint can have either a positive form $[[α̂⁺ :≈ iP]]$ or
a negative form $[[α̂⁻ :≈ iN]]$, however, the right-hand side of the constraint
cannot contain algorithmic type variables.
The set of constraints is denoted as $[[SC]]$. 
We assume that the constraints in $[[SC]]$ restrict distinct variables.

We separately define $[[UC]]$ as a set consisting of unification constraints only.
This is done to simplify the representation of the algorithm. 
The unification, which we use as a subroutine of the subtyping algorithm,
can only produce unification constraints.
A set of unification constraints can be resolved in a simpler way than
a general constraint set. This way, 
the separation of the unification constraint resolution into a separate
procedure allows us to better decompose the structure of the algorithm,
and thus, simplify the inductive proofs.

@@

  \begin{figure}[h]
    \begin{center}
      $[[Θ]] ::= \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$
    \end{center}
    \label{fig:constraint-context}
    \caption{Constraint Contexts}
  \end{figure}

@@

\paragraph{Quantifiers}  
Rules \ruleref{\ottdruleAForallLabel} and 
\ruleref{\ottdruleAExistsLabel} are symmetric. 
According to the declarative specification, 
the quantified variables on the left-hand side must be instantiated 
with types, which, however, are not known in advance.
We deal with it by algorithmization (\cref{todo})
of the quantified variables:
we introduce fresh algorithmic variables
$[[puas]]$ or $[[nuas]]$,
put them into the constraint context $[[Θ]]$
(specifying that they must be instantiated in the extended context
$[[Γ, pbs]]$ or $[[Γ, nbs]]$) and substitute the quantified variables
for them in the input type.

After that, the algorithm proceeds with the recursive call, returning
constraints $[[SC]]$. As the output, the algorithm removes the freshly
introduced algorithmic variables from the constraint context, This operation is
sound: it is guaranteed that $[[SC]]$ always has a solution, but the specific
instantiation of the freshly introduced algorithmic variables is not important,
as they do not occur in the input types.

@@

\paragraph{Algorithmic Variable}
If one of the sides of the subtyping is a unification variable,
the algorithm must create a new restriction. 
Because the right-hand side of the subtyping is always declarative,
it is only the left-hand side that can be a unification variable.
Moreover, another invariant we preserve prevents the negative
algorithmic variables from occurring in types during the
 negative subtyping algorithm. It means
that the only possible form of the subtyping here is $[[α̂⁺]] [[≥]] [[iP]]$,
which is covered by \ruleref{\ottdruleAPUVarLabel}.

@@

The merge procedure that we discuss in this section, allows one to 
combine two constraint sets into one. It might seem that 
just taking the union of the two sets would be enough, however,
we require the constraint sets to have a certain structure,
in particular not to have two entries restricting the same algorithmic
 variable---we call such entries \emph{matching}.
The matching entries must be combined into one constraint entry, that 
would represent their conjunction. 
This way, to merge two constraint sets, we unite the entries of two
sets, and then merge the matching pairs.

@@

Suppose that one of the input entries, say $[[scE1]]$, is a unification 
constraint entry. Then the resulting entry $[[scE1]]$ must coincide with it 
(up-to-equivalence), and thus, it is only required to check that $[[scE2]]$ 
is implied by $[[scE1]]$.
\begin{itemize}
  \item If $[[scE2]]$ is also a restricting entry, then the types on the right-hand side
    of $[[scE1]]$ and $[[scE2]]$ must be equivalent,
    as given by rules \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}.
  \item If $[[scE2]]$ is a supertype restriction $[[α̂⁺ :≥ iP]]$,
    the algorithm must check that the type assigned by $[[scE1]]$ is a supertype of $[[iP]]$.
    The corresponding symmetric rules are \ruleref{\ottdruleSCMESupEqLabel} and \ruleref{\ottdruleSCMEEqSupLabel}.
\end{itemize}

If both input entries are supertype restrictions: $[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$,
then their conjunction is $[[α̂⁺ :≥ iP ∨ iQ]]$, as given by \ruleref{\ottdruleSCMESupSupLabel}.
The least upper bound---$[[iP ∨ iQ]]$ is the least supertype of both $[[iP]]$ and $[[iQ]]$,
and this way, $[[α̂⁺ :≥ iP ∨ iQ]]$ is the weakest constraint entry that implies
$[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$. The algorithm for finding the least upper bound
is discussed in \cref{todo}.

@@

\paragraph{The Type Upgarde}
The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
It exploits the idea that the free variables of a positive type $[[iQ]]$
cannot disappear in its subtypes. It means that if 
a type $[[iP]]$ has free variables not occurring 
in $[[iP']]$, then any common supertype of $[[iP]]$
and $[[iP']]$ must not contain these variables either.
This way, any supertype of $[[iP]]$
not containing certain variables $[[pnas]]$ must also be 
a supertype of $[[iP' = [pnbs/pnas]iP ]]$, where $[[pnbs]]$ are fresh;
and vice versa: any common supertype of $[[iP]]$ and $[[iP']]$
does not contain $[[pnas]]$ nor $[[pnbs]]$.

>>>>

In this section, we present the algorithmization of the declarative system described above.
The algorithmic system follows the structure of the declarative specification closely.
First, it is also given by a set of inference rules, which, however,
must be mode-correct (\cite{dunfield2020:bidirectional}), \ie
the output of each rule is always uniquely defined by its input.
And second, most of the declarative rules 
(except for the rules \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel})
have a unique algorithmic counterpart, 
which simplifies reasoning about the algorithm and its correctness proofs.

@@

\begin{figure}[h]
  \begin{multicols}{2}
      Negative Algorithmic Variables\\
      $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, \dots\\
      \columnbreak
      Positive Algorithmic Variables\\
      $[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, \dots\\
  \end{multicols}
  \hfill\\
  \begin{multicols}{2}
      \ottuNShort
      \columnbreak
      \ottuPShort
  \end{multicols}
  \hfill\\
  Algorithmic Type Context\\
   $[[Ξ]] = \{[[α1̂±]], \dots, [[αn̂±]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
  \hfill\\
  Constraint Type Context\\
   $[[Θ]] ::= \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
  \caption{Algorithmic Syntax}
  \label{fig:algo-syntax}
\end{figure}

\paragraph{Algorithmic Variables}
To design a mode-correct inference system, we slightly modify the language we operate on.
The entities (terms, types, contexts) that the algorithm manipulates we call \emph{algorithmic}. 
They extend the previously defined declarative terms and types by adding 
\emph{algorithmic type variables} (\aka unification variables). 
The algorithmic variables represent unknown types, 
which cannot be inferred immediately but are promised to be instantiated
as the algorithm proceeds.

We denote algorithmic variables as $[[α̂⁺]]$, $[[β̂⁻]]$, \dots to distinguish them from
normal variables $[[α⁺]]$, $[[β⁻]]$.
In a few places, we replace the quantified variables $[[pas]]$ 
with their algorithmic counterpart $[[puas]]$.
The procedure of replacing declarative variables with algorithmic ones we call 
\emph{algorithmization} and denote as $[[ nuas/nas ]]$ and $[[ puas/pas ]]$.

\paragraph{Algorithmic Types}
The syntax of algorithmic types extends the declarative syntax by adding
algorithmic variables as new terminals. We add positive algorithmic variables $[[α̂⁺]]$ 
to the syntax of positive types, and negative algorithmic variables $[[α̂⁻]]$ to the 
syntax of negative types. All the constructors of the system can be applied 
to \emph{algorithmic} types, however, algorithmic variables cannot be abstracted by the
quantifiers $[[∀]]$ and $[[∃]]$.

\paragraph{Algorithmic Contexts and Well-formedness}
To specify when algorithmic types are well-formed, 
we define algorithmic contexts $[[Ξ]]$ as sets of algorithmic variables.
Then $[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness
judgment of algorithmic terms defined as expected. 
Informally, they check that all free declarative variables are in $[[Γ]]$, 
and all free algorithmic variables are in $[[Ξ]]$.
In addition to the rules repeating the declarative definition \cref{todo}, 
we have two base cases for the algorithmic variables:
 \ruleref{\ottdruleWFATPUVarLabel} and \ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf})

\begin{figure}[h]
\begin{multicols}{2}
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleWFATPUVarLabeled{}}
  \columnbreak\\
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleWFATNUVarLabeled{}}
\end{multicols}
\label{fig:algo-wf}
\caption{Well-formedness of Algorithmic Types}
\end{figure}

\paragraph{Algorithmic Normalization}
Similarly to well-formedness, the normalization of algorithmic types is defined
by extending the declarative definition with the algorithmic variables.
To the rules repeating the declarative normalization, we add
rules saying that normalization is trivial on algorithmic variables (see \cref{fig:algo-nf}).

@@

\subsection{Type Constraints}
As the algorithm proceeds, it accumulates the information 
about the algorithmic type variables in the form of \emph{constraints}.
In our system, the constraints can be of two kinds: 
\emph{subtyping constraints} and \emph{unification constraints}.
The subtyping constraint can only have a positive shape $[[α̂⁺ :≥ iP]]$, \ie it 
restricts a positive algorithmic variable to be a 
supertype of a certain declarative type---this is one of the invariants that we 
preserve in the algorithm.
The unification constraint can have either a positive form $[[α̂⁺ :≈ iP]]$ or
a negative form $[[α̂⁻ :≈ iN]]$, however, the right-hand side of the constraint
cannot contain algorithmic type variables.
The set of constraints is denoted as $[[SC]]$. 
We assume that each algorithmic variable can be restricted by at most one constraint.

We separately define $[[UC]]$ as a set consisting of unification constraints only.
This is done to simplify the representation of the algorithm. 
The unification algorithm, which we use as a subroutine of the subtyping algorithm,
can only produce unification constraints.
A set of unification constraints can be resolved in a simpler way than
a general constraint set. This way, 
the separation of the unification constraint resolution into a separate
procedure allows us to better decompose the structure of the algorithm,
and thus, simplify the inductive proofs.

@@

@@

\paragraph{Quantifiers}  
Rules \ruleref{\ottdruleAForallLabel} and 
\ruleref{\ottdruleAExistsLabel} are symmetric. 
Declaratively, the quantified variables on the left-hand side must 
be instantiated with types, which are not known beforehand.
We address this problem by algorithmization (\cref{todo})
of the quantified variables.
The rule introduces fresh algorithmic variables
$[[puas]]$ or $[[nuas]]$,
puts them into the constraint context $[[Θ]]$
(specifying that they must be instantiated in the extended context
$[[Γ, pbs]]$ or $[[Γ, nbs]]$) and substitute the quantified variables
for them in the input type.

After algorithmization of the quantified variables, 
the algorithm proceeds with the recursive call, returning
constraints $[[SC]]$. As the output, the algorithm removes the freshly
introduced algorithmic variables from the constraint context, This operation is
sound: it is guaranteed that $[[SC]]$ always has a solution, but the specific
instantiation of the freshly introduced algorithmic variables is not important,
as they do not occur in the input types.

@@

\paragraph{Algorithmic Variable}
If one of the sides of the subtyping is a unification variable,
the algorithm must create a new constraint. 
Because the right-hand side of the subtyping is always declarative,
it is only the left-hand side that can be a unification variable.
Moreover, another invariant we preserve prevents the negative
algorithmic variables from occurring in types during the
 negative subtyping algorithm. It means
that the only possible form of the subtyping here is $[[α̂⁺]] [[≥]] [[iP]]$,
which is covered by \ruleref{\ottdruleAPUVarLabel}.

@@

In this section, we discuss the constraint merging procedure.
It allows one to combine two constraint sets into one. 
A simple union of two constraint sets is not sufficient, 
since the resulting set must not contain two entries restricting the 
same algorithmic variable---we call such entries \emph{matching}.
The matching entries must be combined into \emph{one} constraint entry, that 
would represent their conjunction. 
This way, to merge two constraint sets, we unite the entries of two
sets, and then merge the matching pairs.

@@

Suppose that one of the input entries, say $[[scE1]]$, is a unification 
constraint entry. Then the resulting entry $[[scE1]]$ must coincide with it 
(up-to-equivalence), and thus, it is only required to check that $[[scE2]]$ 
is implied by $[[scE1]]$.
\begin{itemize}
  \item If $[[scE2]]$ is also a restricting entry, then the types on the right-hand side
    of $[[scE1]]$ and $[[scE2]]$ must be equivalent,
    as given by rules \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}.
  \item If $[[scE2]]$ is a supertype constraint $[[α̂⁺ :≥ iP]]$,
    the algorithm must check that the type assigned by $[[scE1]]$ is a supertype of $[[iP]]$.
    The corresponding symmetric rules are \ruleref{\ottdruleSCMESupEqLabel} and \ruleref{\ottdruleSCMEEqSupLabel}.
\end{itemize}

If both input entries are supertype constraints: $[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$,
then their conjunction is $[[α̂⁺ :≥ iP ∨ iQ]]$, as given by \ruleref{\ottdruleSCMESupSupLabel}.
The least upper bound---$[[iP ∨ iQ]]$ is the least supertype of both $[[iP]]$ and $[[iQ]]$,
and this way, $[[α̂⁺ :≥ iP ∨ iQ]]$ is the weakest constraint entry that implies
$[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$. The algorithm for finding the least upper bound
is discussed in \cref{todo}.

@@

\paragraph{The Type Upgarde}
The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
It exploits the idea that the free variables of a positive type $[[iQ]]$
cannot disappear in its subtypes (see \cref{prop:subtyping-preserves-fv}). 
It means that if 
a type $[[iP]]$ has free variables not occurring 
in $[[iP']]$, then any common supertype of $[[iP]]$
and $[[iP']]$ must not contain these variables either.
This way, any supertype of $[[iP]]$
not containing certain variables $[[pnas]]$ must also be 
a supertype of $[[iP' = [pnbs/pnas]iP ]]$, where $[[pnbs]]$ are fresh;
and vice versa: any common supertype of $[[iP]]$ and $[[iP']]$
does not contain $[[pnas]]$ nor $[[pnbs]]$.
